---
title: "conformity_data_analysis"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, error = F, message = F, warning = F)

```

```{r libraries, include=FALSE, message=FALSE, warning=FALSE}

# package for convenience functions (e.g. ggplot2, dplyr, etc.)
library(tidyverse)

# package for Bayesian regression
library(brms)

library(rstan)

library(bridgesampling)

# parallel execution of Stan code
options(mc.cores = parallel::detectCores())

# nicer global knitr options
knitr::opts_chunk$set(warning = FALSE, message = FALSE, 
                      cache = TRUE, fig.align = 'center')
set.seed(1702)
```

# Analysis
```{r}
# loads the cleaned dataset
cleaned_data <- read_csv('../../Data/01_pilot/02_cleaned_pilot_data.csv')
# creates the dataset used for the stan models
stan_data <- as.list(c(cleaned_data, N = dim(cleaned_data)[1]))
```


```{r}
# creates the stan model for the alternative hypothesis
write("
data{
    int<lower=1> N;
    int dilemma_rating[N];
    int both_infos[N];
    int ingroup_norm[N];
}
parameters{
    ordered[5] cutpoints;
    real<lower=0> bIn;
    real bBoth;
}
transformed parameters{
    real bOut = -1.4166666667 * bIn;
}
model{
    vector[N] phi;
    cutpoints ~ normal( 0 , 10 );
    bIn ~ normal( 0.816 , 0.5 );
    bBoth ~ normal( 0 , 0.5 );
    for ( i in 1:N ) {
        phi[i] = bIn * ingroup_norm[i] + bBoth * both_infos[i] + bOut * both_infos[i] * ingroup_norm[i];
        dilemma_rating[i] ~ ordered_logistic( phi[i] , cutpoints );
    }
}
generated quantities{
    vector[N] phi;
    real dev;
    vector[N] log_lik;
    dev = 0;
    for ( i in 1:N ) {
        phi[i] = bIn * ingroup_norm[i] + bBoth * both_infos[i] + bOut * both_infos[i] * ingroup_norm[i];
    dev = dev + (-2)*ordered_logistic_lpmf( dilemma_rating[i] | phi[i] , cutpoints );
    log_lik[i] = ordered_logistic_lpmf( dilemma_rating[i] | phi[i] , cutpoints );
    }
}", "../stan_models/01_alternative_model.stan")

alternative_model <- "../stan_models/01_alternative_model.stan"
```

```{r}
# fits the stan model for the alternative hypothesis
stanfit_alt <- stan(file = alternative_model, data = stan_data, iter = 10000, chains = 4,  warmup = 500, seed =123, control=list(adapt_delta = 0.99))
# feeds the fitted stan model back into brms
brms_fit_alt <- brm(dilemma_rating ~ ingroup_norm * both_infos, data = cleaned_data, family = cumulative('logit'), fit = stanfit_alt, save_all_pars = TRUE)
```

```{r}
# creates the stan model for sct model
write("
data{
    int<lower=1> N;
    int dilemma_rating[N];
    int both_infos[N];
    int ingroup_norm[N];
    int ingroup_agree[N];
    int outgroup_disagree[N];
}
parameters{
    real<lower=0> bIn;
    real bBoth;
    real<lower=0> bOut;
    ordered[5] cutpoints;
}
model{
    vector[N] phi;
    cutpoints ~ normal( 0 , 10 );
    bBoth ~ normal( 0 , 0.5 );
    bOut ~ normal( 0 , 0.5 );
    bIn ~ normal( 0.6/0.75 * 1.02 , 0.5 );
    for ( i in 1:N ) {
        phi[i] = bIn * ingroup_norm[i] * ingroup_agree[i] + bBoth * both_infos[i] + bOut * both_infos[i] * ingroup_norm[i] * outgroup_disagree[i];
        dilemma_rating[i] ~ ordered_logistic( phi[i] , cutpoints );
    }
}
generated quantities{
    vector[N] phi;
    real dev;
    vector[N] log_lik;
    dev = 0;
    for ( i in 1:N ) {
        phi[i] = bIn * ingroup_norm[i] * ingroup_agree[i] + bBoth * both_infos[i] + bOut * both_infos[i] * ingroup_norm[i] * outgroup_disagree[i];
        dev = dev + (-2)*ordered_logistic_lpmf( dilemma_rating[i] | phi[i] , cutpoints );
        log_lik[i] = ordered_logistic_lpmf( dilemma_rating[i] | phi[i] , cutpoints );
    }
}", "../stan_models/02_sct_model.stan")

sct_model <- "../stan_models/02_sct_model.stan"
```

```{r}
# fits the stan model for the self-categorization theory (sct)
stanfit_sct <- stan(file = sct_model, data = stan_data, iter = 10000, chains = 4,  warmup = 500, seed =123, control=list(adapt_delta = 0.99))

# feeds the fitted stan model back into brms
brms_fit_sct <- brm(dilemma_rating ~ ingroup_norm * ingroup_agree + both_infos + ingroup_norm * both_infos * outgroup_disagree, data = cleaned_data, family = cumulative('logit'), fit = stanfit_sct, save_all_pars = TRUE)
```


```{r}
# computes the marginal likelihoods
marg_lik_sct <- brms::bridge_sampler(samples = brms_fit_sct)
marg_lik_alt <- brms::bridge_sampler(samples = brms_fit_alt)

# computes the bayes_factors (1st in favor of the alternative model and 2nd in favor of the sct model)
bf(marg_lik_alt,marg_lik_sct)
bf(marg_lik_sct,marg_lik_alt)
```


# Exploratory Analysis

## Checking for Possible Main Effect of the Type of Dilemma (dilemma_version)

### Preparing Data for the Extending the Models with a dilemma_version Term
```{r}
explor_data <- cleaned_data %>% mutate(
  dilemma_type = case_when(
    dilemma_type == 'philosophical_dilemma' ~ 1,
    dilemma_type == 'real_dilemma' ~ -1
  ),
  dilemma_type = as.integer(dilemma_type)
)
explor_data

explor_stan_data <- as.list(c(explor_data, N = dim(explor_data)[1]))
```


### Extending the Alternative Model with a Dilemma Version Main-Effect

```{r}

# creates the stan model for the alternative hypothesis
write("
data{
    int<lower=1> N;
    int dilemma_rating[N];
    int both_infos[N];
    int ingroup_norm[N];
    int dilemma_type[N];
}
parameters{
    ordered[5] cutpoints;
    real<lower=0> bIn;
    real bBoth;
    real bTyp;
}
transformed parameters{
    real bOut = -1.4166666667 * bIn;
}
model{
    vector[N] phi;
    cutpoints ~ normal( 0 , 10 );
    bIn ~ normal( 0.816 , 0.5 );
    bBoth ~ normal( 0 , 0.5 );
    bTyp ~ normal(0, 3);
    for ( i in 1:N ) {
        phi[i] = bIn * ingroup_norm[i] + bBoth * both_infos[i] + bOut * both_infos[i] * ingroup_norm[i] + bTyp * dilemma_type[i];
        dilemma_rating[i] ~ ordered_logistic( phi[i] , cutpoints );
    }
}
generated quantities{
    vector[N] phi;
    real dev;
    vector[N] log_lik;
    dev = 0;
    for ( i in 1:N ) {
        phi[i] = bIn * ingroup_norm[i] + bBoth * both_infos[i] + bOut * both_infos[i] * ingroup_norm[i] + bTyp * dilemma_type[i];
    dev = dev + (-2)*ordered_logistic_lpmf( dilemma_rating[i] | phi[i] , cutpoints );
    log_lik[i] = ordered_logistic_lpmf( dilemma_rating[i] | phi[i] , cutpoints );
    }
}", "../stan_models/03_explor_alternative_model.stan")

explor_alternative_model <- "../stan_models/03_explor_alternative_model.stan"
```

```{r}
explor_stanfit_alt <- stan(file = explor_alternative_model, data = explor_stan_data, iter = 10000, chains = 4,  warmup = 500, seed =123, control=list(adapt_delta = 0.99))

explor_brms_fit_alt <- brm(dilemma_rating ~ dilemma_type + ingroup_norm * both_infos, data = explor_data, family = cumulative('logit'), fit = explor_stanfit_alt, save_all_pars = TRUE)
```

#### Does inlcuding the dilemma_version term helps the alternative model to better explain the data?
```{r}
marg_lik_explor_alt <- brms::bridge_sampler(samples = explor_brms_fit_alt)

bf(marg_lik_explor_alt,marg_lik_alt)
```

#### What direction does a possible dilemma_version effect have in the alternative model?
```{r}
summary(explor_brms_fit_alt)
```

### Extending the SCT Model with a Dilemma Version Main-Effect

```{r}
write("
data{
    int<lower=1> N;
    int dilemma_rating[N];
    int both_infos[N];
    int ingroup_norm[N];
    int ingroup_agree[N];
    int outgroup_disagree[N];
    int dilemma_type[N];
}
parameters{
    real<lower=0> bIn;
    real bBoth;
    real<lower=0> bOut;
    real bTyp;
    ordered[5] cutpoints;
}
model{
    vector[N] phi;
    cutpoints ~ normal( 0 , 10 );
    bBoth ~ normal( 0 , 0.5 );
    bOut ~ normal( 0 , 0.5 );
    bIn ~ normal( 0.6/0.75 * 1.02 , 0.5 );
    bTyp ~ normal(0, 3);
    for ( i in 1:N ) {
        phi[i] = bIn * ingroup_norm[i] * ingroup_agree[i] + bBoth * both_infos[i] + bOut * both_infos[i] * ingroup_norm[i] * outgroup_disagree[i] + bTyp * dilemma_type[i];
        dilemma_rating[i] ~ ordered_logistic( phi[i] , cutpoints );
    }
}
generated quantities{
    vector[N] phi;
    real dev;
    vector[N] log_lik;
    dev = 0;
    for ( i in 1:N ) {
        phi[i] = bIn * ingroup_norm[i] * ingroup_agree[i] + bBoth * both_infos[i] + bOut * both_infos[i] * ingroup_norm[i] * outgroup_disagree[i] + bTyp * dilemma_type[i];
        dev = dev + (-2)*ordered_logistic_lpmf( dilemma_rating[i] | phi[i] , cutpoints );
        log_lik[i] = ordered_logistic_lpmf( dilemma_rating[i] | phi[i] , cutpoints );
    }
}", "../stan_models/04_explor_sct_model.stan")

explor_sct_model <- "../stan_models/04_explor_sct_model.stan"
```

```{r}
explor_stanfit_sct <- stan(file = explor_sct_model, data = explor_stan_data, iter = 10000, chains = 4,  warmup = 500, seed =123, control=list(adapt_delta = 0.99))

explor_brms_fit_sct <- brm(dilemma_rating ~ ingroup_norm * ingroup_agree + both_infos + ingroup_norm * both_infos * outgroup_disagree + dilemma_type, data = explor_data, family = cumulative('logit'), fit = explor_stanfit_sct, save_all_pars = TRUE)
```

#### Does inlcuding the dilemma_version term helps the sct-model to better explain the data?
```{r}
marg_lik_explor_sct <- brms::bridge_sampler(samples = explor_brms_fit_sct)

bf(marg_lik_explor_sct,marg_lik_sct)
```

#### What direction does a possible dilemma_version effect have in the sct-model?
```{r}
summary(explor_brms_fit_sct)
```

## Investigating Performance of Competing Models for each Dilemma Type

### Splitting the dataset via the dilemma_version variable and preparing the new datasets for the stan models
```{r}
real_data <- cleaned_data %>% filter(dilemma_type == 'real_dilemma')
philo_data <- cleaned_data %>% filter(dilemma_type == 'philosophical_dilemma')

real_stan_data <- as.list(c(real_data, N = dim(real_data)[1]))
philo_stan_data <- as.list(c(philo_data, N = dim(philo_data)[1]))
```

### Fitting the alternative models
```{r}
real_stanfit_alt <- stan(file = alternative_model, data = real_stan_data, iter = 10000, chains = 4,  warmup = 500, seed =123, control=list(adapt_delta = 0.99))
philo_stanfit_alt <- stan(file = alternative_model, data = philo_stan_data, iter = 10000, chains = 4,  warmup = 500, seed =123, control=list(adapt_delta = 0.99))


real_brms_fit_alt <- brm(dilemma_rating ~ ingroup_norm * both_infos, data = real_data, family = cumulative('logit'), fit = real_stanfit_alt, save_all_pars = TRUE)
philo_brms_fit_alt <- brm(dilemma_rating ~ ingroup_norm * both_infos, data = philo_data, family = cumulative('logit'), fit = philo_stanfit_alt, save_all_pars = TRUE)
```

### Fitting the sct models
```{r}
real_stanfit_sct <- stan(file = sct_model, data = real_stan_data, iter = 10000, chains = 4,  warmup = 500, seed =123, control=list(adapt_delta = 0.99))
philo_stanfit_sct <- stan(file = sct_model, data = philo_stan_data, iter = 10000, chains = 4,  warmup = 500, seed =123, control=list(adapt_delta = 0.99))


real_brms_fit_sct <- brm(dilemma_rating ~ ingroup_norm * ingroup_agree + both_infos + ingroup_norm * both_infos * outgroup_disagree, data = real_data, family = cumulative('logit'), fit = real_stanfit_sct, save_all_pars = TRUE)
philo_brms_fit_sct <- brm(dilemma_rating ~ ingroup_norm * ingroup_agree + both_infos + ingroup_norm * both_infos * outgroup_disagree, data = philo_data, family = cumulative('logit'), fit = philo_stanfit_sct, save_all_pars = TRUE)
```

### Comparing the models for the realistic dilemma decisions
```{r}
real_marg_lik_sct <- brms::bridge_sampler(samples = real_brms_fit_sct)
real_marg_lik_alt <- brms::bridge_sampler(samples = real_brms_fit_alt)

bf(real_marg_lik_alt,real_marg_lik_sct)
bf(real_marg_lik_sct,real_marg_lik_alt)
```

### Comparing the models for the philosophical dilemma decisions
```{r}
philo_marg_lik_sct <- brms::bridge_sampler(samples = philo_brms_fit_sct)
philo_marg_lik_alt <- brms::bridge_sampler(samples = philo_brms_fit_alt)

bf(philo_marg_lik_alt,philo_marg_lik_sct)
bf(philo_marg_lik_sct,philo_marg_lik_alt)
```

