---
title: "conformity_data_analysis"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, error = F, message = F, warning = F)

```

```{r libraries, include=FALSE, message=FALSE, warning=FALSE}

# package for convenience functions (e.g. ggplot2, dplyr, etc.)
library(tidyverse)

# package for this course
#library(aida)

# package for Bayesian regression
library(brms)

library(rstan)

library(bridgesampling)

# parallel execution of Stan code
options(mc.cores = parallel::detectCores())

# nicer global knitr options
knitr::opts_chunk$set(warning = FALSE, message = FALSE, 
                      cache = TRUE, fig.align = 'center')
set.seed(1702)
```


```{r}
# reads the raw experimental data
raw_data <- read_csv('data_pilot.csv')
```
# Demographics

```{r}
# creates an overview of the demographics of the participants
# CAUTION: Not mandatory for participants to enter these informations (therefore na values possible)
demographics <- raw_data %>% filter(trial_name == 'topic_choice') %>%
  summarize(N = length(submission_id),
            num_provided_age = sum(!is.na(age)),
            mean_age = mean(age, na.rm = TRUE),
            min_age = min(age, na.rm = TRUE),
            max_age = max(age, na.rm = TRUE),
            sd_age = sd(age, na.rm = TRUE),
            median_ag = median(age, na.rm = TRUE),
            num_provided_gender = sum(!is.na(gender)),
            num_males = sum(gender=="Männlich", na.rm = TRUE),
            num_females = sum(gender=="Weiblich", na.rm = TRUE),
            num_divers = sum(gender=="Divers", na.rm = TRUE),
            num_provided_education = sum(!is.na(education)),
            num_lower_degree = sum(education == "Real- oder Hauptschulabschluss", na.rm = TRUE),
            num_high_school_degree = sum(education == 'Abitur', na.rm = TRUE),
            num_higher_degree = sum(education == 'Universitärer Abschluss', na.rm = TRUE),
            num_provided_language = sum(!is.na(languages)),
            num_german_native_speakers = sum(languages == 'Deutsch', na.rm = TRUE),
            num_other_native_speakers = sum(languages == 'Andere', na.rm = TRUE)
  )

demographics
```


# Cleaning data

```{r}
tidy_data <- raw_data %>% 
  # removes all trials in which the participant just presses a next button and all test submissions
  filter(response != 'next', languages != "TEST") %>%
  # creates a trials variables that contains a descriptions of the trial in the shape trialName_trialNumber
  mutate(trials = paste(trial_name, trial_number, sep = "_")) %>%
  # removes variables unnecessary for further cleaning or analysis
  select(-QUD, -RT, - startDate, - startTime, -age, -gender, -education, -question, -option1, -option2, -trial_name, -trial_number, -experiment_id) %>%
  # changes the form so each trial gets its own column with the response as value
  pivot_wider(
   names_from = trials,
   values_from = response
) %>%
  # renames variables for better understanding
  rename(ingroup_norm = ingroup_info_a, real_dilemma = dilemma_decision1_4, philosophical_dilemma = dilemma_decision2_4, feeling_real_dilemma = dilemma_decision1_5, feeling_philosophical_dilemma = dilemma_decision2_5) %>%
  # checks whether all 3 understanding checks were answered correctly
  mutate(
      passed_understanding_check = case_when(
      understanding_check_1 == "Es ist irrelevant, wie lange du für deine Antworten benötigt hast."&
      understanding_check_2 == "Anfangs wurden dir mehrere Themen zur Auswahl gestellt." &
      understanding_check_3 == "Wir können dir leider keine Aufwandsentschädigung oder VP Stunden anbieten, sind dir aber sehr dankbar." ~ TRUE)) %>%
  # removes the answers to the understanding checks (information if passed now contained in passed_understanding_check)
  select(-understanding_check_1, -understanding_check_2, -understanding_check_3) %>%
  # creates ingroup_daily and ingroup_philo variables that contain the ingroup norm in the rows of their dilemma version
  mutate(
    ingroup_real = case_when(
      !is.na(real_dilemma) ~ ingroup_norm
    ),
    ingroup_philo = case_when(
      !is.na(philosophical_dilemma) ~ ingroup_norm
    )) %>%
  # removes the ingroup_norm variable (information now stored in ingroup_philo and ingroup_real)
  select(-ingroup_norm) %>%
  # replaces na values by the existing value of the respective variable in other rows of the subject
  group_by(submission_id) %>%
  mutate(
    both_infos = unique(both_infos[!is.na(both_infos)]),
    topic_choice_1 = unique(topic_choice_1[!is.na(topic_choice_1)]),
    statement_rating_1 = unique(statement_rating_1[!is.na(statement_rating_1)]),
    philosophical_dilemma = unique(philosophical_dilemma[!is.na(philosophical_dilemma)]),
    feeling_philosophical_dilemma = unique(feeling_philosophical_dilemma[!is.na(feeling_philosophical_dilemma)]),
    real_dilemma = unique(real_dilemma[!is.na(real_dilemma)]),
    feeling_real_dilemma = unique(feeling_real_dilemma[!is.na(feeling_real_dilemma)]),
    group_ident_1 = unique(group_ident_1[!is.na(group_ident_1)]),
    group_ident_2 = unique(group_ident_2[!is.na(group_ident_2)]),
    passed_understanding_check = unique(passed_understanding_check[!is.na(passed_understanding_check)]),
    ingroup_real = unique(ingroup_real[!is.na(ingroup_real)]),
    ingroup_philo = unique(ingroup_philo[!is.na(ingroup_philo)])
    ) %>%
  ungroup() %>% 
  # now the rows for each participant contain the same information, so they can be merged to obtain 1 row per participant
  unique() %>%
  # creates the ingroup_norm for both realistic and philosophical dilemma according to the coding of the original study: -1 if ingroup prefers active action (e.g. changing the direction of the train) and 1 if ingroup prefers passive action (e.g. not changing the direction of the train)
  mutate(
    ingroup_norm_real = case_when(
      ingroup_real == TRUE ~ -1,
      ingroup_real == FALSE ~ 1,
    ),
    ingroup_norm_philo = case_when(
      ingroup_philo == TRUE ~ -1,
      ingroup_philo == FALSE ~ 1,
    )) %>%
  # the variables used for creating the ingroup_norm variable in the previous step can now be dropped (info is contained in ingroup_norm)
  select(-ingroup_real, -ingroup_philo) %>%
  # change all logical variables to numeric variables (TRUE ~ 1 and FALSE ~ 0)
  mutate_if(is.logical,as.numeric) %>%
  # changes format so that there are two rows per participant (one for each dilemma) and creates the variable dilemma_version (denoting the type of dilemma) and dilemma_rating (the participant's decision for that dilemma)
  pivot_longer(
    cols = c(6, 8),
    names_to = "dilemma_type",
    values_to = "dilemma_rating"
  ) %>%
  mutate(
    # now that we have one row per dilemma, the two ingroup_norm variables (daily and philo) can be reduced to one ingroup_norm variable
    ingroup_norm = case_when(
      dilemma_type == "philosophical_dilemma" ~ ingroup_norm_philo,
      dilemma_type == "real_dilemma" ~ ingroup_norm_real
    ),
    # creates variable that indicates whether the subject identifies with its ingroup
    ingroup_agree = case_when(
      statement_rating_1 <= 0 & group_ident_2 > 0 ~ 1,
      statement_rating_1 >= 0 & group_ident_1 > 0 ~ 1,
      statement_rating_1 <= 0 & group_ident_2 <= 0 ~ 0,
      statement_rating_1 >= 0 & group_ident_1 <= 0 ~ 0,
    ),
    # create variable that indicates whether the subject does not identify with its outgroup (again: statement_rating_1 = 0 not necessary)
    outgroup_disagree = case_when(
      statement_rating_1 <= 0 & group_ident_1 < 0 ~ 1,
      statement_rating_1 >= 0 & group_ident_2 < 0 ~ 1,
      statement_rating_1 <= 0 & group_ident_1 >= 0 ~ 0,
      statement_rating_1 >= 0 & group_ident_2 >= 0 ~ 0,
    )
  ) %>%
  # removes ingroup_norm_daily and ingroup_norm_philo used for creating the one ingroup_norm variable
   select(-ingroup_norm_real, -ingroup_norm_philo) %>%
  # corrects the type of integer variables
  mutate(
      submission_id = as.integer(submission_id),
      dilemma_rating = as.integer(dilemma_rating),
      both_infos = as.integer(both_infos),
      ingroup_norm = as.integer(ingroup_norm),
      ingroup_agree = as.integer(ingroup_agree),
      outgroup_disagree = as.integer(outgroup_disagree),
      statement_rating_1 = as.integer(statement_rating_1),
      topic_choice_1 = as.integer(topic_choice_1),
      group_ident_1 = as.integer(group_ident_1),
      group_ident_2 = as.integer(group_ident_2),
      feeling_real_dilemma = as.integer(feeling_real_dilemma),
      feeling_philosophical_dilemma = as.integer(feeling_philosophical_dilemma),
      passed_understanding_check = as.integer(passed_understanding_check)
    )


tidy_data
```

```{r}
# counts all participants that indicated that there native language is not German
language_summary <- tidy_data %>% filter(dilemma_type == 'real_dilemma') %>% 
  summarise(
    not_german_native_speaker = sum(languages == 'Andere')
  )
language_summary
# removes those participants that are not German native speakers
tidy_data <- tidy_data %>% filter(languages != 'Andere')

# creates a summary of how many participant were exluded for either or both reasons
exclusion_summary <- tidy_data %>% filter(dilemma_type == 'real_dilemma') %>%
  summarise(
    failed_understanding_check = sum(passed_understanding_check == 0 & !between(statement_rating_1, -2, 2)),
    too_low_statement_rating = sum(between(statement_rating_1, -2, 2) & passed_understanding_check != 0),
    both = sum(passed_understanding_check == 0 & between(statement_rating_1, -2, 2)),
    total = failed_understanding_check + too_low_statement_rating + both
  )

exclusion_summary
```

```{r}
# removes any participant with a statement_rating_1 between (including) -2 and 2 and all that failed the understanding check
# additionally selects the variables relevant for the analysis
cleaned_data <- tidy_data %>% 
  filter(!between(statement_rating_1, -2, 2)) %>%
  filter(passed_understanding_check == TRUE) %>% 
  select(submission_id, dilemma_rating, both_infos, ingroup_norm, ingroup_agree, outgroup_disagree, dilemma_version)

cleaned_data
```

# Analysis
```{r}
# creates the dataset used for the stan models
stan_data <- as.list(c(cleaned_data, N = dim(cleaned_data)[1]))
```


```{r}
# creates the stan model for the alternative hypothesis
write("
data{
    int<lower=1> N;
    int dilemma_rating[N];
    int both_infos[N];
    int ingroup_norm[N];
}
parameters{
    ordered[5] cutpoints;
    real<lower=0> bIn;
    real bBoth;
}
transformed parameters{
    real bOut = -1.4166666667 * bIn;
}
model{
    vector[N] phi;
    cutpoints ~ normal( 0 , 10 );
    bIn ~ normal( 0.816 , 0.5 );
    bBoth ~ normal( 0 , 0.5 );
    for ( i in 1:N ) {
        phi[i] = bIn * ingroup_norm[i] + bBoth * both_infos[i] + bOut * both_infos[i] * ingroup_norm[i];
        dilemma_rating[i] ~ ordered_logistic( phi[i] , cutpoints );
    }
}
generated quantities{
    vector[N] phi;
    real dev;
    vector[N] log_lik;
    dev = 0;
    for ( i in 1:N ) {
        phi[i] = bIn * ingroup_norm[i] + bBoth * both_infos[i] + bOut * both_infos[i] * ingroup_norm[i];
    dev = dev + (-2)*ordered_logistic_lpmf( dilemma_rating[i] | phi[i] , cutpoints );
    log_lik[i] = ordered_logistic_lpmf( dilemma_rating[i] | phi[i] , cutpoints );
    }
}", "alternative_model.stan")

alternative_model <- "alternative_model.stan"
```

```{r}
# fits the stan model for the alternative hypothesis
stanfit_alt <- stan(file = alternative_model, data = stan_data, iter = 10000, chains = 4,  warmup = 500, seed =123, control=list(adapt_delta = 0.99))
# feeds the fitted stan model back into brms
brms_fit_alt <- brm(dilemma_rating ~ ingroup_norm * both_infos, data = cleaned_data, family = cumulative('logit'), fit = stanfit_alt, save_all_pars = TRUE)
```

```{r}
# creates the stan model for sct model
write("
data{
    int<lower=1> N;
    int dilemma_rating[N];
    int both_infos[N];
    int ingroup_norm[N];
    int ingroup_agree[N];
    int outgroup_disagree[N];
}
parameters{
    real<lower=0> bIn;
    real bBoth;
    real<lower=0> bOut;
    ordered[5] cutpoints;
}
model{
    vector[N] phi;
    cutpoints ~ normal( 0 , 10 );
    bBoth ~ normal( 0 , 0.5 );
    bOut ~ normal( 0 , 0.5 );
    bIn ~ normal( 0.6/0.75 * 1.02 , 0.5 );
    for ( i in 1:N ) {
        phi[i] = bIn * ingroup_norm[i] * ingroup_agree[i] + bBoth * both_infos[i] + bOut * both_infos[i] * ingroup_norm[i] * outgroup_disagree[i];
        dilemma_rating[i] ~ ordered_logistic( phi[i] , cutpoints );
    }
}
generated quantities{
    vector[N] phi;
    real dev;
    vector[N] log_lik;
    dev = 0;
    for ( i in 1:N ) {
        phi[i] = bIn * ingroup_norm[i] * ingroup_agree[i] + bBoth * both_infos[i] + bOut * both_infos[i] * ingroup_norm[i] * outgroup_disagree[i];
        dev = dev + (-2)*ordered_logistic_lpmf( dilemma_rating[i] | phi[i] , cutpoints );
        log_lik[i] = ordered_logistic_lpmf( dilemma_rating[i] | phi[i] , cutpoints );
    }
}", "sct_model.stan")

sct_model <- "sct_model.stan"
```

```{r}
# fits the stan model for the self-categorization theory (sct)
stanfit_sct <- stan(file = sct_model, data = stan_data, iter = 10000, chains = 4,  warmup = 500, seed =123, control=list(adapt_delta = 0.99))

# feeds the fitted stan model back into brms
brms_fit_sct <- brm(dilemma_rating ~ ingroup_norm * ingroup_agree + both_infos + ingroup_norm * both_infos * outgroup_disagree, data = cleaned_data, family = cumulative('logit'), fit = stanfit_sct, save_all_pars = TRUE)
```


```{r}
# computes the marginal likelihoods
marg_lik_sct <- brms::bridge_sampler(samples = brms_fit_sct)
marg_lik_alt <- brms::bridge_sampler(samples = brms_fit_alt)

# computes the bayes_factors (1st in favor of the alternative model and 2nd in favor of the sct model)
bf(marg_lik_alt,marg_lik_sct)
bf(marg_lik_sct,marg_lik_alt)
```

```{r}
bf(marg_lik_alt, marg_lik_alt)
```

```{r}
bf(marg_lik_sct, marg_lik_sct)
```
# Exploratory Analysis

## Checking for Possible Main Effect of the Type of Dilemma (dilemma_version)

### Preparing Data for the Extending the Models with a dilemma_version Term
```{r}
explor_data <- cleaned_data %>% mutate(
  dilemma_type = case_when(
    dilemma_type == 'philosophical_dilemma' ~ 1,
    dilemma_type == 'real_dilemma' ~ -1
  ),
  dilemma_type = as.integer(dilemma_type)
)
explor_data

explor_stan_data <- as.list(c(explor_data, N = dim(explor_data)[1]))
```


### Extending the Alternative Model with a Dilemma Version Main-Effect

```{r}

# creates the stan model for the alternative hypothesis
write("
data{
    int<lower=1> N;
    int dilemma_rating[N];
    int both_infos[N];
    int ingroup_norm[N];
    int dilemma_type[N];
}
parameters{
    ordered[5] cutpoints;
    real<lower=0> bIn;
    real bBoth;
    real bTyp;
}
transformed parameters{
    real bOut = -1.4166666667 * bIn;
}
model{
    vector[N] phi;
    cutpoints ~ normal( 0 , 10 );
    bIn ~ normal( 0.816 , 0.5 );
    bBoth ~ normal( 0 , 0.5 );
    bTyp ~ normal(0, 3);
    for ( i in 1:N ) {
        phi[i] = bIn * ingroup_norm[i] + bBoth * both_infos[i] + bOut * both_infos[i] * ingroup_norm[i] + bTyp * dilemma_type[i];
        dilemma_rating[i] ~ ordered_logistic( phi[i] , cutpoints );
    }
}
generated quantities{
    vector[N] phi;
    real dev;
    vector[N] log_lik;
    dev = 0;
    for ( i in 1:N ) {
        phi[i] = bIn * ingroup_norm[i] + bBoth * both_infos[i] + bOut * both_infos[i] * ingroup_norm[i] + bTyp * dilemma_type[i];
    dev = dev + (-2)*ordered_logistic_lpmf( dilemma_rating[i] | phi[i] , cutpoints );
    log_lik[i] = ordered_logistic_lpmf( dilemma_rating[i] | phi[i] , cutpoints );
    }
}", "explor_alternative_model.stan")

explor_alternative_model <- "explor_alternative_model.stan"
```

```{r}
explor_stanfit_alt <- stan(file = explor_alternative_model, data = explor_stan_data, iter = 10000, chains = 4,  warmup = 500, seed =123, control=list(adapt_delta = 0.99))

explor_brms_fit_alt <- brm(dilemma_rating ~ dilemma_type + ingroup_norm * both_infos, data = explor_data, family = cumulative('logit'), fit = explor_stanfit_alt, save_all_pars = TRUE)
```

#### Does inlcuding the dilemma_version term helps the alternative model to better explain the data?
```{r}
marg_lik_explor_alt <- brms::bridge_sampler(samples = explor_brms_fit_alt)

bf(marg_lik_explor_alt,marg_lik_alt)
```

#### What direction does a possible dilemma_version effect have in the alternative model?
```{r}
summary(explor_brms_fit_alt)
```

### Extending the SCT Model with a Dilemma Version Main-Effect

```{r}
write("
data{
    int<lower=1> N;
    int dilemma_rating[N];
    int both_infos[N];
    int ingroup_norm[N];
    int ingroup_agree[N];
    int outgroup_disagree[N];
    int dilemma_type[N];
}
parameters{
    real<lower=0> bIn;
    real bBoth;
    real<lower=0> bOut;
    real bTyp;
    ordered[5] cutpoints;
}
model{
    vector[N] phi;
    cutpoints ~ normal( 0 , 10 );
    bBoth ~ normal( 0 , 0.5 );
    bOut ~ normal( 0 , 0.5 );
    bIn ~ normal( 0.6/0.75 * 1.02 , 0.5 );
    bTyp ~ normal(0, 3);
    for ( i in 1:N ) {
        phi[i] = bIn * ingroup_norm[i] * ingroup_agree[i] + bBoth * both_infos[i] + bOut * both_infos[i] * ingroup_norm[i] * outgroup_disagree[i] + bTyp * dilemma_type[i];
        dilemma_rating[i] ~ ordered_logistic( phi[i] , cutpoints );
    }
}
generated quantities{
    vector[N] phi;
    real dev;
    vector[N] log_lik;
    dev = 0;
    for ( i in 1:N ) {
        phi[i] = bIn * ingroup_norm[i] * ingroup_agree[i] + bBoth * both_infos[i] + bOut * both_infos[i] * ingroup_norm[i] * outgroup_disagree[i] + bTyp * dilemma_type[i];
        dev = dev + (-2)*ordered_logistic_lpmf( dilemma_rating[i] | phi[i] , cutpoints );
        log_lik[i] = ordered_logistic_lpmf( dilemma_rating[i] | phi[i] , cutpoints );
    }
}", "explor_sct_model.stan")

explor_sct_model <- "explor_sct_model.stan"
```

```{r}
explor_stanfit_sct <- stan(file = explor_sct_model, data = explor_stan_data, iter = 10000, chains = 4,  warmup = 500, seed =123, control=list(adapt_delta = 0.99))

explor_brms_fit_sct <- brm(dilemma_rating ~ ingroup_norm * ingroup_agree + both_infos + ingroup_norm * both_infos * outgroup_disagree + dilemma_type, data = explor_data, family = cumulative('logit'), fit = explor_stanfit_sct, save_all_pars = TRUE)
```

#### Does inlcuding the dilemma_version term helps the sct-model to better explain the data?
```{r}
marg_lik_explor_sct <- brms::bridge_sampler(samples = explor_brms_fit_sct)

bf(marg_lik_explor_sct,marg_lik_sct)
```

#### What direction does a possible dilemma_version effect have in the sct-model?
```{r}
summary(explor_brms_fit_sct)
```

## Investigating Performance of Competing Models for each Dilemma Type

### Splitting the dataset via the dilemma_version variable and preparing the new datasets for the stan models
```{r}
real_data <- cleaned_data %>% filter(dilemma_type == 'real_dilemma')
philo_data <- cleaned_data %>% filter(dilemma_type == 'philosophical_dilemma')

real_stan_data <- as.list(c(real_data, N = dim(real_data)[1]))
philo_stan_data <- as.list(c(philo_data, N = dim(philo_data)[1]))
```

### Fitting the alternative models
```{r}
real_stanfit_alt <- stan(file = alternative_model, data = real_stan_data, iter = 10000, chains = 4,  warmup = 500, seed =123, control=list(adapt_delta = 0.99))
philo_stanfit_alt <- stan(file = alternative_model, data = philo_stan_data, iter = 10000, chains = 4,  warmup = 500, seed =123, control=list(adapt_delta = 0.99))


real_brms_fit_alt <- brm(dilemma_rating ~ ingroup_norm * both_infos, data = real_data, family = cumulative('logit'), fit = real_stanfit_alt, save_all_pars = TRUE)
philo_brms_fit_alt <- brm(dilemma_rating ~ ingroup_norm * both_infos, data = philo_data, family = cumulative('logit'), fit = philo_stanfit_alt, save_all_pars = TRUE)
```

### Fitting the sct models
```{r}
real_stanfit_sct <- stan(file = sct_model, data = real_stan_data, iter = 10000, chains = 4,  warmup = 500, seed =123, control=list(adapt_delta = 0.99))
philo_stanfit_sct <- stan(file = sct_model, data = philo_stan_data, iter = 10000, chains = 4,  warmup = 500, seed =123, control=list(adapt_delta = 0.99))


real_brms_fit_sct <- brm(dilemma_rating ~ ingroup_norm * ingroup_agree + both_infos + ingroup_norm * both_infos * outgroup_disagree, data = real_data, family = cumulative('logit'), fit = real_stanfit_sct, save_all_pars = TRUE)
philo_brms_fit_sct <- brm(dilemma_rating ~ ingroup_norm * ingroup_agree + both_infos + ingroup_norm * both_infos * outgroup_disagree, data = philo_data, family = cumulative('logit'), fit = philo_stanfit_sct, save_all_pars = TRUE)
```

### Comparing the models for the realistic dilemma decisions
```{r}
real_marg_lik_sct <- brms::bridge_sampler(samples = real_brms_fit_sct)
real_marg_lik_alt <- brms::bridge_sampler(samples = real_brms_fit_alt)

bf(real_marg_lik_alt,real_marg_lik_sct)
bf(real_marg_lik_sct,real_marg_lik_alt)
```

### Comparing the models for the philosophical dilemma decisions
```{r}
philo_marg_lik_sct <- brms::bridge_sampler(samples = philo_brms_fit_sct)
philo_marg_lik_alt <- brms::bridge_sampler(samples = philo_brms_fit_alt)

bf(philo_marg_lik_alt,philo_marg_lik_sct)
bf(philo_marg_lik_sct,philo_marg_lik_alt)
```

